Supermartingales and Termination Analysis of Statistical PCF

### 2020-12-28 LICS21/antitone.tex
For consistency: global replace indicator function 1_{...} -> {\bf 1}_{...}
Changed S and \mu to \entrosp and \mu_\entrosp for all files except confluent.tex

\begin{definition}
We say that a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
\[
\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot 1_{\set{T > n}}
\]
\end{definition}

@@

Assume for a contradiction that $\mathbb P[T > B_x] > x$.
Then, take $n \leq B_x$. 
We have
%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
\begin{calculation}
\expect{Y_n-Y_{n+1}}
  \step[=]{$\calF_n \subseteq \calF_{n+1}$, def.~\& linearity of cond.~expectation}
\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
  \step[\geq]{antitone strict assumption}
\expect{\epsilon(Y_n) \cdot 1_{\set{T > n}}}
%  \step[=]{definition of conditional expectation for a discrete variable}
\step[=]{def.~expectation conditioning on an \emph{event}}
\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
  \step[\geq]{Jensen's inequality}
\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
  \step[\geq]{proved earlier}
\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
\end{calculation}
Therefore, by a telescoping sum
\begin{align*}
\expect{Y_{B_x}} 
&= \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})\\
&\leq -1 < 0 
\end{align*}
%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
\end{proof}

@@

\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
If a closed SPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
\end{theorem}
\begin{proof}
As in %Theorem \ref{thm:basicRankingSoundness}, 
\Cref{thm:rankable implies termination},
take the set $S$ of traces \lo{TODO: We should find a better name for traces $s \in S$.} as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$. As before, $(Y_n)_{n \geq 0}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$. Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY} < \infty$ a.s.
\end{proof}

@@

  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$. 
  The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions. Define $f_2(N) = \int_S f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu(\mathrm d s)$. For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone partial ranking functions. In order to show that $f_2$ is an antitone partial ranking function, it therefore suffices to show that the value that $f_2$ would have had at each term if $f$ were not defined at that term is at least the expectation of $f_2$ after one reduction step (plus $\epsilon(f_2(N))$ if the reduction step is a $tY$-reduction). For any term $N$ which is not of the form $E[R]$ for some $\tY$-redex $R$, this is trivial. If $R$ is a $\tY$-redex, then $\epsilon(f_2(N)) \leq \int_S \epsilon(f_1(\nnext(N,s)))\, \mu(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded. Therefore, the (total) function $f_2$, which agrees with $f$ on $f$'s domain, is an antitone ranking function on $M$ (with the same function $\epsilon$ if it's convex).
\end{proof}

@@

Although the stronger induction hypothesis was necessary for the induction to work, the only reason this is needed is to prove that $y_n > 0$ for all $n$, which implies that $Y_n \geq 0$. 
The other condition for $(Y_n)$ to be an antitone-strict supermartingale is
\begin{align*}
    & \expect{Y_{n+1} \mid \mathcal F_n} \\
    &=  \expect{y_{n+1} \, 1_{\set{T \geq n+1}} \mid \mathcal F_n} \\
    &=  y_{n+1} \, \expect{1_{\set{T \geq n+1}} \mid \mathcal F_n} \\ 
    &=  y_{n+1} \, 1_{\set{T \geq n}} \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
    &=  (y_n - \epsilon(y_n)) \, 1_{\set{T \geq n}} \\
    &= Y_n - \epsilon(y_n) \, 1_{\set{T \geq n}}
\end{align*}
therefore $(Y_n)_n$ is an antitone strict supermartingale with respect to $(T,\epsilon)$.
\end{proof}

>>>>

\begin{definition}
We say that a supermartingale $(Y_n)_{n \geq 0}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \geq 0}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
\[
\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot {\bf 1}_{\set{T > n}}
\]
\end{definition}

@@

Assume for a contradiction that $\mathbb P[T > B_x] > x$.
Then, take $n \leq B_x$. 
We have
%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
\begin{calculation}
\expect{Y_n-Y_{n+1}}
  \step[=]{$\calF_n \subseteq \calF_{n+1}$, def.~\& linearity of cond.~expectation}
\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
  \step[\geq]{antitone strict assumption}
\expect{\epsilon(Y_n) \cdot {\bf 1}_{\set{T > n}}}
%  \step[=]{definition of conditional expectation for a discrete variable}
\step[=]{def.~expectation conditioning on an \emph{event}}
\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
  \step[\geq]{Jensen's inequality}
\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
  \step[\geq]{proved earlier}
\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
\end{calculation}
Therefore, by a telescoping sum
\begin{align*}
\expect{Y_{B_x}} 
&= \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})\\
&\leq -1 < 0 
\end{align*}
%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
\end{proof}

@@

\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
If a closed SPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
\end{theorem}
\begin{proof}
As in %Theorem \ref{thm:basicRankingSoundness}, 
\Cref{thm:rankable implies termination},
take the set $\entrosp$ of traces \lo{TODO: We should find a better name for traces $s \in \entrosp$.} as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$. As before, $(Y_n)_{n \geq 0}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$. Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY} < \infty$ a.s.
\end{proof}

@@

  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$. 
  The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions. Define $f_2(N) = \int_\entrosp f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu_{\entrosp}(\mathrm d s)$. For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone partial ranking functions. In order to show that $f_2$ is an antitone partial ranking function, it therefore suffices to show that the value that $f_2$ would have had at each term if $f$ were not defined at that term is at least the expectation of $f_2$ after one reduction step (plus $\epsilon(f_2(N))$ if the reduction step is a $tY$-reduction). For any term $N$ which is not of the form $E[R]$ for some $\tY$-redex $R$, this is trivial. If $R$ is a $\tY$-redex, then $\epsilon(f_2(N)) \leq \int_\entrosp \epsilon(f_1(\nnext(N,s)))\, \mu_{\entrosp}(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded. Therefore, the (total) function $f_2$, which agrees with $f$ on $f$'s domain, is an antitone ranking function on $M$ (with the same function $\epsilon$ if it's convex).
\end{proof}

@@

Although the stronger induction hypothesis was necessary for the induction to work, the only reason this is needed is to prove that $y_n > 0$ for all $n$, which implies that $Y_n \geq 0$. 
The other condition for $(Y_n)$ to be an antitone-strict supermartingale is
\begin{align*}
    & \expect{Y_{n+1} \mid \mathcal F_n} \\
    &=  \expect{y_{n+1} \, {\bf 1}_{\set{T \geq n+1}} \mid \mathcal F_n} \\
    &=  y_{n+1} \, \expect{{\bf 1}_{\set{T \geq n+1}} \mid \mathcal F_n} \\ 
    &=  y_{n+1} \, {\bf 1}_{\set{T \geq n}} \frac{\mathbb P[T > n+1]}{\mathbb P[T > n]} \\
    &=  (y_n - \epsilon(y_n)) \, {\bf 1}_{\set{T \geq n}} \\
    &= Y_n - \epsilon(y_n) \, {\bf 1}_{\set{T \geq n}}
\end{align*}
therefore $(Y_n)_n$ is an antitone strict supermartingale with respect to $(T,\epsilon)$.
\end{proof}

### 2020-12-28 LICS21/supermartingales.tex
Changed S and \mu to \entrosp and \mu_\entrosp for all files except confluent.tex

%\emph{Notation.}
\changed[lo]{Henceforth, fix the probability space $(S, \calF, \mu_{\entrosp})$, and a closed SPCF term $M$.
For $n \geq 0$, define the random variables $M_n(s) := \pi_0 (\red^n(M, s))$, and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.}

@@

Given an $\omega$-sequence (e.g.~$s \in \entrosp$) and $m \geq 0$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
%Fix a closed SPCF term $M \in \Lambda^0$, and a ranking function $f$ for it, satisfying $f(N) = 0$ iff $N$ is a value.
%Let $n \geq 0$, define the following random variables on the probability space $(S, \calF, \mu)$:
%\begin{align*}
%M_n(s) &:= \pi_0 (\red^n(M, s))\\
%\ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|,
%\end{align*}
For $n \geq 0$, define
\begin{align*}
%M_n(s) &:= \pi_0 (\red^n(M, s))\\
\ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|
\end{align*}
so that $\pi_1 (\red^n(M, s)) = %\overbrace{\pi_t( \cdots (\pi_t}^{\ndraw n s}(s))$. 
\pi_t( \cdots (\pi_t (s))$, with $\pi_t$ applied ${\ndraw n s}$ times.
%and the filtration $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \citep{DBLP:conf/icfp/BorgstromLGS16}.
%$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$). 
\iffalse
\akr{$\#\mathrm{draw}_n$ is not a stopping time. It doesn't look like you actually use this claim anyway, so it should just be fine to remove, but did you mean something different?} \lo{I agree, and I don't actually use this claim.}
\fi
Take $s \in A \in \calF_n$ with $\ndraw{n}{s} = l$.
For any $s'\in \entrosp$, if $s_{\leq l} = s_{\leq l}'$ then $s' \in A$.
It follows that ${\set{s_{\leq l}} \cdot I^\omega} \subseteq A$.
\lo{NOTE. To fix notation:
\(
\sigma(M_{n}) = \sigma\big(\set{M_{n}^{-1}(\alpha_j, U_{\alpha_j})
\mid \alpha_j\in \mathsf{Sk}_j, U_{\alpha_j} \in \mathcal{B}(\Real^j), j \geq 0}\big)
\)
}
%Take $M \in \Lambda^0$ (closed SPCF terms). Define, for each $n \in \omega$, the random variable $M_n : S \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.

@@

\begin{proof}
We show the non-trivial case of $i = 2$.
First we express 
\begin{equation}
f(M_{n+1})[s \in \mathbf{T}_2] = \sum_{i \in \calI} 
f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i]
\label{eqn:f 2 n+1}
\end{equation}
where 
\begin{itemize}
\item $\calI$ is a countable indexing set
\item $E_i[\cdot][\tsample] \in \mathsf{Sk}_{j_i}$, and $\sigma_i : \entrosp \to \Real^{j_i}$, and $\rho(s) := \pi_h(\pi_1(\red^n(M, s))) = \pi_h(\pi_t^{l_i}(s)) \in \Real$ 
\item $\set{U_i}_{i \in \calI}$ is a partition of $\mathbf{T}_2$, 
where each $U_i$ is determined by a skeletal term $E_i[\cdot][\tsample]$ and a number (of draws) $l_i \leq n$, satisfying
\(
s \in U_i
\iff
(E_i[\underline{r}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s),
\;
\hbox{for some }r \in \Real^{j_i}.
\)
(We use the (measurable) function $\sigma_i : S \to \Real^{j_i}$ to skolemise the existentially quantified $r$.)
Equivalently
\[
U_i := \#\mathrm{draw}_n^{-1}(l_i) \cap M_n^{-1}(E_i[\cdot][\tsample], \Real^{j_i}) \in \calF_n.
\]
%\lo{I.e.~$(E_i[\underline{\sigma_i(s)}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s)$ iff $s \in U_i$.}
\end{itemize}

@@

Let $f$ be a ranking function on $M$ satisfying $f(N) = 0$ iff $N$ is a value.
Define random variables on the probability space $(S, \calF, \mu)$:
\begin{align}
T_{-1}(s) & := -1 \nonumber \\
T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value} \nonumber\\
& \qquad\qquad\qquad\qquad\qquad \hbox{or of form } E[\tY \lambda x. N] \} \nonumber \\
X_n(s) & := f(M_n(s)) \nonumber \\
Y_n(s) & := X_{T_n}(s) \nonumber \\
T_M^\tY(s) &:= \min \set{n \mid M_{T_n(s)}(s) \textrm{ is a value}} \label{eq:y stopping time} 
%T_M(s) & := \min \set{n \mid X_n(s) = 0} \nonumber
\end{align}

>>>>

%\emph{Notation.}
\changed[lo]{Henceforth, fix the probability space $(\entrosp, \calF, \mu_{\entrosp})$, and a closed SPCF term $M$.
For $n \geq 0$, define the random variables $M_n(s) := \pi_0 (\red^n(M, s))$, and the filtration $(\mathcal{F}_n)_{n \geq 0}$ where $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.}

@@

Given an $\omega$-sequence (e.g.~$s \in \entrosp$) and $m \geq 0$, we write $s_{\leq m} \in I^m$ to mean the prefix of $s$ of length $m$.
%Fix a closed SPCF term $M \in \Lambda^0$, and a ranking function $f$ for it, satisfying $f(N) = 0$ iff $N$ is a value.
%Let $n \geq 0$, define the following random variables on the probability space $(S, \calF, \mu)$:
%\begin{align*}
%M_n(s) &:= \pi_0 (\red^n(M, s))\\
%\ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|,
%\end{align*}
For $n \geq 0$, define
\begin{align*}
%M_n(s) &:= \pi_0 (\red^n(M, s))\\
\ndraw{n}{s} &:= |\{k < n \mid \exists E: M_k(s) = E[\tsample]\}|
\end{align*}
so that $\pi_1 (\red^n(M, s)) = %\overbrace{\pi_t( \cdots (\pi_t}^{\ndraw n s}(s))$. 
\pi_t( \cdots (\pi_t (s))$, with $\pi_t$ applied ${\ndraw n s}$ times.
%and the filtration $\mathcal{F}_n := \sigma(M_1, \cdots, M_n)$.
The $\calF_n$-measurability of $M_n$ (and hence of $\#\mathrm{draw}_n$) follows from \citep{DBLP:conf/icfp/BorgstromLGS16}.
%$\#\mathrm{draw}_n$ is a stopping time (bounded by $n$). 
\iffalse
\akr{$\#\mathrm{draw}_n$ is not a stopping time. It doesn't look like you actually use this claim anyway, so it should just be fine to remove, but did you mean something different?} \lo{I agree, and I don't actually use this claim.}
\fi
Take $s \in A \in \calF_n$ with $\ndraw{n}{s} = l$.
For any $s'\in \entrosp$, if $s_{\leq l} = s_{\leq l}'$ then $s' \in A$.
It follows that ${\set{s_{\leq l}} \cdot I^\omega} \subseteq A$.
\lo{NOTE. To fix notation:
\(
\sigma(M_{n}) = \sigma\big(\set{M_{n}^{-1}(\alpha_j, U_{\alpha_j})
\mid \alpha_j\in \mathsf{Sk}_j, U_{\alpha_j} \in \mathcal{B}(\Real^j), j \geq 0}\big)
\)
}
%Take $M \in \Lambda^0$ (closed SPCF terms). Define, for each $n \in \omega$, the random variable $M_n : \entrosp \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.

@@

\begin{proof}
We show the non-trivial case of $i = 2$.
First we express 
\begin{equation}
f(M_{n+1})[s \in \mathbf{T}_2] = \sum_{i \in \calI} 
f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i]
\label{eqn:f 2 n+1}
\end{equation}
where 
\begin{itemize}
\item $\calI$ is a countable indexing set
\item $E_i[\cdot][\tsample] \in \mathsf{Sk}_{j_i}$, and $\sigma_i : \entrosp \to \Real^{j_i}$, and $\rho(s) := \pi_h(\pi_1(\red^n(M, s))) = \pi_h(\pi_t^{l_i}(s)) \in \Real$ 
\item $\set{U_i}_{i \in \calI}$ is a partition of $\mathbf{T}_2$, 
where each $U_i$ is determined by a skeletal term $E_i[\cdot][\tsample]$ and a number (of draws) $l_i \leq n$, satisfying
\(
s \in U_i
\iff
(E_i[\underline{r}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s),
\;
\hbox{for some }r \in \Real^{j_i}.
\)
(We use the (measurable) function $\sigma_i : \entrosp \to \Real^{j_i}$ to skolemise the existentially quantified $r$.)
Equivalently
\[
U_i := \#\mathrm{draw}_n^{-1}(l_i) \cap M_n^{-1}(E_i[\cdot][\tsample], \Real^{j_i}) \in \calF_n.
\]
%\lo{I.e.~$(E_i[\underline{\sigma_i(s)}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s)$ iff $s \in U_i$.}
\end{itemize}

@@

Let $f$ be a ranking function on $M$ satisfying $f(N) = 0$ iff $N$ is a value.
Define random variables on the probability space $(\entrosp, \calF, \mu)$:
\begin{align}
T_{-1}(s) & := -1 \nonumber \\
T_{n+1}(s) & := \min \{ k \mid k>T_n(s), M_k(s) \textrm{ a value} \nonumber\\
& \qquad\qquad\qquad\qquad\qquad \hbox{or of form } E[\tY \lambda x. N] \} \nonumber \\
X_n(s) & := f(M_n(s)) \nonumber \\
Y_n(s) & := X_{T_n}(s) \nonumber \\
T_M^\tY(s) &:= \min \set{n \mid M_{T_n(s)}(s) \textrm{ is a value}} \label{eq:y stopping time} 
%T_M(s) & := \min \set{n \mid X_n(s) = 0} \nonumber
\end{align}
